### 2012-10-12 paper_sean/hyperneat.tex
made every change as requested by Jeff, except for p from Wilcoxon rank sum test

% !TEX root =main.tex
HyperNEAT is an algorithm for evolving artificial neural networks (ANNs)~\cite{stanley2009hypercube}. It has been repeatedly described in detail~\cite{stanley2009hypercube, gauci2007generating, clune2011performance},  so here we provide only a summary. Instead of directly encoding the weights of the ANN individually on the genome, in HyperNEAT the genome is a compositional pattern producing network (CPPN)~\cite{stanley2007compositional}. The CPPN genome specifies the weights in a similar way to how natural organisms develop. In nature, phenotypic attributes are specified as a function of their geometric location, and such positional information is conveyed through chemical morphogen gradients~\cite{carroll2005endless}. For example, the concentration of one chemical could indicate the position along the anterior-posterior (head-to-tail) axis, and another chemical in bands could indicate if a cell is in an odd-numbered or even-numbered segment. Based on the relative concentrations of these chemicals, a cell can know where it is geometrically and, thus, what type of cell to become~\cite{carroll2005endless}.

With CPPNs this process is abstracted as a network of math functions that operate in a Cartesian geometric space. The coordinates of phenotypic elements are provided as inputs to the CPPN and the output specifies a phenotypic trait. For example, when CPPNs encode 2D pictures, the coordinates of each pixel are iteratively input into the genome and the output is the grayscale value at that coordinate~\cite{secretan2011picbreeder}. Because a CPPN network is composed of math functions, those functions can create geometric regularities in the phenotype. For example, a Gaussian function of the x-axis can provide left-right symmetry, and a repeating function (e.g. sine) of the y-axis could provide a related motif such as segmentation~(Fig.~\figref{hyperneat_bug_example}). Both 2D pictures and 3D objects evolved with CPPNs look like natural and engineered objects, and contain complex, regularities like symmetries and repeated motifs, with and without variation~\cite{secretan2011picbreeder, clune2011objects}.

@@

In HyperNEAT, CPPNs encode the weights of the connections between neurons as a function of the geometric locations of those neurons~(\figref{hyperneatExplanation}). The Cartesian coordinates of the two neurons at the end of each connection are input into the CPPN, and the output is the weight of that connection. If the output is smaller than a threshold, the weight is set to zero, functionally removing the connection. The process is repeated for each possible connection. Just as in 2D pictures and 3D objects, the CPPN can create complex, regular, geometric patterns (e.g. left-right symmetry or repeated modules), but in this case the patterns are in the weights of the neural network~\cite{clune2011performance, }. The neural regularities produced by HyperNEAT enable significantly improved performance on problems that are regular~\cite{clune2011performance, stanley2009hypercube}, including evolving quadruped gaits~\cite{clune2011performance, clune2009evolving, clune2009sensitivity}.

>>>>

% !TEX root =main.tex
HyperNEAT is an algorithm for evolving artificial neural networks (ANNs)~\cite{stanley2009hypercube}. It has been repeatedly described in detail~\cite{stanley2009hypercube,gauci2007generating,clune2011performance},  so here we provide only a summary. Instead of directly encoding the weights of the ANN individually on the genome, in HyperNEAT the genome is a compositional pattern producing network (CPPN)~\cite{stanley2007CPPN}. The CPPN genome specifies the weights in a similar way to how natural organisms develop. In nature, phenotypic attributes are specified as a function of their geometric location, and such positional information is conveyed through chemical morphogen gradients~\cite{carroll2005endless}. For example, the concentration of one chemical could indicate the position along the anterior-posterior (head-to-tail) axis, and another chemical in bands could indicate if a cell is in an odd-numbered or even-numbered segment. Based on the relative concentrations of these chemicals, a cell can know where it is geometrically and, thus, what type of cell to become~\cite{carroll2005endless}.

With CPPNs this process is abstracted as a network of math functions that operate in a Cartesian geometric space. The coordinates of phenotypic elements are provided as inputs to the CPPN and the output specifies a phenotypic trait. For example, when CPPNs encode 2D pictures, the coordinates of each pixel are iteratively input into the genome and the output is the grayscale value at that coordinate~\cite{secretan2011picbreeder}. Because a CPPN network is composed of math functions, those functions can create geometric regularities in the phenotype. For example, a Gaussian function of the x-axis can provide left-right symmetry, and a repeating function (e.g. sine) of the y-axis could provide a related motif such as segmentation~(\figref{hyperneat_bug_example}). Both 2D pictures and 3D objects evolved with CPPNs look like natural and engineered objects, and contain complex, regularities like symmetries and repeated motifs, with and without variation~\cite{secretan2011picbreeder,clune2011objects}.

@@

In HyperNEAT, CPPNs encode the weights of the connections between neurons as a function of the geometric locations of those neurons~(\figref{hyperneatExplanation}). The Cartesian coordinates of the two neurons at the end of each connection are input into the CPPN, and the output is the weight of that connection. If the output is smaller than a threshold, the weight is set to zero, functionally removing the connection. The process is repeated for each possible connection. Just as in 2D pictures and 3D objects, the CPPN can create complex, regular, geometric patterns (e.g. left-right symmetry or repeated modules), but in this case the patterns are in the weights of the neural network~\cite{clune2011performance}. The neural regularities produced by HyperNEAT enable significantly improved performance on problems that are regular~\cite{clune2011performance,stanley2009hypercube}, including evolving quadruped gaits~\cite{clune2011performance,clune2009evolving,clune2009sensitivity}.

### 2012-10-12 paper_sean/results.tex
changed a few things, including table..and cut out some of my comments that sean has addressed
made every change as requested by Jeff, except for p from Wilcoxon rank sum test

% !TEX root =main.tex
Because of the noise inherent in all physical robots, especially with different copies of robots, the best way to compare two algorithms is in simulation. The gaits evolved with HyperNEAT in simulation significantly outperformed the gaits evolved with a GA by Glette et al. 2012~(Table~\ref{resultsTable}). INSERT-RES-OF-SIM-PARAGRAPH.

Gaits evolved in the simulator with HyperNEAT and then transferred to reality (HyperNEAT+Simulator) outperformed all previous gaits reported for the QuadraTot robot platform~(Table~\ref{resultsTable}). Videos of the gaits in simulation and on robot can be viewed at http://MAKEMEACREATIVEMACHINESLINK.
The fastest previously reported gait moved 17.8 cm/s, but was evaluated on a copy of the robot in Norway~\cite{glette}. To control for the possibility that the robot itself may be different, we ran that gait 10 times on our copy of the robot. We found that our robot is inherently slower, and the average speed of the Glette et al. 2012 gait was 12.95 cm/s on our copy of the robot, with a maximum speed of 13.76 cm/s.
The speed of the fastest gait produced in this paper was 14.5 cm/s on our copy of the robot, which is 5.4\% faster than the best gait from Glette et al. 2012, once robot itself is controlled for. This result is even more significant when one considers that the gait produced by HyperNEAT with a simulator evolved with 20000 (33\%) fewer evaluations per run than Glette et al. 2010, a decision mandated by the available computational resources.

@@

HyperNEAT+Simulator gaits also outperformed the best gait evolved with HyperNEAT directly in hardware~\cite{yos:clune} by 49.5\%, most likely due to the difference in the population size (200 vs. 9) and the number of generations (200 vs. 20), leading to a total difference in the number of evaluations between the two studies of 40000 vs 180~\cite{yos:clune}. Another potential cause of improved performance is that with a simulator HyperNEAT had much less noise in the evaluation process during early generations: it could thus find coordinated, regular, gaits, which are both higher-performing and more likely to transfer. On the physical robot, the noise in the evaluation was substantial, preventing effective learning and the discovery of effective gaits~\cite{yos:clune}. Additionally, we cannot rule out that the addition in this study of a punishment for high-frequency gaits aided performance, especially since in simulation the gaits were high-frequency in a few early generations before rapidly settling at a lower frequency due to the penalty. It is possible that quickly leaving the high-frequency area of the search space caused a significant performance boost.

% table!
\begin{table}
\caption{The number of evaluations per run, simulated (average) and real trial (max) velocities in [cm/s]. Data used from Yosinksi et al. 2011~\cite{yos:clune}, Glette et al. 2012~\cite{glette}, and Shen et al. 2012~\cite{haocheng}.
*17.8 cm/s was recorded on Glette et al. 2012~\cite{glette} but the same gait tried on our robot averaged over ten runs was 12.951 cm/s}  %\tablabel{results}
\begin{center}
\begin{tabular}{|l|r|r|r|r|}
\hline
                                         & Evaluations  & Simulated Velocity  & Real Velocity \\
\hline
Best parameter optimization learning    &153    & --    & 4.6 \\
\hline
Hardware HyperNEAT                  & 180         & --         &   9.7     \\
\hline
Simulated GA              & 60000       & 16.4       &   13.0*     \\
\hline
RL PoWER Spline                          & 300         & --         &   11.1 \\
\hline
Simulated HyperNEAT                      & 40000       & 25.4       &   14.5 \\
\hline
\end{tabular}
\end{center}
\label{resultsTable}
\end{table}
%JMC All numbers in the same column should be to the same precision (e.g. for the real velocity column, one decimal place). Also, can you right-justify the numbers (while still keeping them roughly centered)? Comparing numbers that are not right-justified is hard...so if you can't center and right-justify, then I let's see if just having it be right justified is better.

Qualitatively, the HyperNEAT gaits are regular and coordinated, as has been found in simulation~\cite{clune2009evolving, clune2011performance}. This is evident in the servo position plots where the gaits produced in this study are smooth and are similar between the servos~\figref{servo_plot_111}. This result is important because it confirms that HyperNEAT can produce the important property of regularity in physically realizable applications, instead of just producing those behaviors in simulation. This conclusion was not obvious after HyperNEAT failed to produce substantially coordinated, regular, effective gaits on this robot platform when evolving directly on the hardware~\cite{yos:clune}.

>>>>

% !TEX root =main.tex
Because of the noise inherent in all physical robots, especially with different copies of robots, the best way to compare two algorithms is in simulation. The gaits evolved with HyperNEAT in simulation significantly outperformed the gaits evolved with a GA by Glette et al. 2012~(Table~\ref{resultsTable}). We compared Kyrre et al 2012's best simulated run to the average of our 20 runs and found that our gait is, in average, greater in fitness by 7.67 cm/s, with 2.0 cm/s standard deviation, for every generation of evolution during the runs. Using the same data, we recorded a fitness of 21.0 cm/s with 4.29 cm/s standard deviation as the mean fitness throughout the evolutionary process while Kyrre et al 2012's data recorded a fitness of 13.35 cm/s with 4.06 cm/s standard deviation. These results show that in simulation, HyperNEAT's gaits evolve faster and have higher fitnesses than those of GA's.

Gaits evolved in the simulator with HyperNEAT and then transferred to reality (HyperNEAT+Simulator) outperformed all previous gaits reported for the QuadraTot robot platform~(Table~\ref{resultsTable}).\footnote{Videos of the gaits in simulation and on the robot can be viewed at http://creativemachines.cornell.edu/HyperNEAT-simulation-quadrupedal-gaits}
The fastest previously reported gait moved 17.8 cm/s, but was evaluated on a copy of the robot in Norway~\cite{glette}. To control for the possibility that the robot itself may be different, we ran that gait 10 times on our copy of the robot. We found that our robot is inherently slower, and the average speed of the Glette et al. 2012 gait was 12.95 cm/s on our copy of the robot, with a maximum speed of 13.76 cm/s.
The speed of the fastest gait produced in this paper was 14.5 cm/s on our copy of the robot, which is 5.4\% faster than the best gait from Glette et al. 2012 performed on our robot. This result is even more significant when one considers that the gait produced by HyperNEAT with a simulator evolved with 20000 (33\%) fewer evaluations per run than Glette et al. 2012, a decision mandated by the available computational resources.

@@

HyperNEAT+Simulator gaits also outperformed the best gait evolved with HyperNEAT directly in hardware~\cite{yos:clune} by 49.5\%, most likely due to the difference in the population size (200 vs. 9) and the number of generations (200 vs. 20), leading to a total difference in the number of evaluations between the two studies of 40000 vs 180~\cite{yos:clune}. Another potential cause of improved performance is that with a simulator HyperNEAT had much less noise in the evaluation process during early generations: it could thus find coordinated, regular, gaits, which are both higher-performing and more likely to transfer. On the physical robot, the noise in the evaluation was substantial, preventing effective learning and the discovery of effective gaits~\cite{yos:clune}. Additionally, we cannot rule out that the addition in this study of a punishment for high-frequency gaits aided performance, especially since in simulation the gaits were high-frequency in a few early generations before rapidly settling at a lower frequency due to the penalty~(\figref{avg_fit_vs_gen}). It is possible that quickly leaving the high-frequency area of the search space caused a significant performance boost.

% table!
\begin{table}
\caption{The number of evaluations per run, simulated (average) and real trial (max) velocities in [cm/s]. Data used from Yosinksi et al. 2011~\cite{yos:clune}, Glette et al. 2012~\cite{glette}, and Shen et al. 2012~\cite{haocheng}.
*17.8 cm/s was recorded on Glette et al. 2012~\cite{glette} but the same gait tried on our robot averaged over ten runs was 12.951 cm/s}  %\tablabel{results}
\begin{center}
\begin{tabular}{|l|r|r|r|r|}
\hline
                                         & Evaluations  & Simulated Velocity  & Real Velocity \\
\hline
Parameterized gaits + optimization    &153    & --    & 4.6 \\
\hline
HyperNEAT in hardware                  & 180         & --         &   9.7     \\
\hline
RL PoWER Spline                          & 300         & --         &   11.1 \\
\hline
GA + simulator              & 60000       & 16.4       &   *13.0     \\
\hline
HyperNEAT + simulator                      & 40000       & 25.4       &   14.5 \\
\hline
\end{tabular}
\end{center}
\label{resultsTable}
\end{table}

Qualitatively, the HyperNEAT gaits are regular and coordinated, as has been found in simulation~\cite{clune2009evolving,clune2011performance}. This is evident in the servo position plots where the gaits produced in this study are smooth and are similar between the servos~(\figref{servo_plot_111}). This result is important because it confirms that HyperNEAT can produce the important property of regularity in physically realizable applications, instead of just producing those behaviors in simulation. This conclusion was not obvious after HyperNEAT failed to produce substantially coordinated, regular, effective gaits on this robot platform when evolving directly on the hardware~\cite{yos:clune}.

### 2012-10-12 paper_sean/simulator.tex
JMC pass through...first forage as far as the distant plains of results. cowered in fear at the conclusion and camped for the night to gather courage
big commit before meeting Jeff/Jason today. Changes: - Shorten - Citation problems - Figure updates (fonts larger) - Figure caption updates - Table right justify - Table sigfigs - Table specified whether max or avg - Adding Kyrre's data to authorlist
made every change as requested by Jeff, except for p from Wilcoxon rank sum test

% !TEX root =main.tex
The simulator\footnote{Available for Linux and Windows at http://quadratot.yosinski.com} was the same one used in Glette et al. \cite{glette}. C++ code for the the Nvidia PhysX physics engine approximately describes the QuadraTot, including the mass and size of the robot components and its degrees of freedom. In the simulator, each individual joint range was limit as described above, but Smart Cropping was not included because it hindered performance by limiting the types of gaits evolution could explore in early generations.

>>>>

% !TEX root =main.tex
The simulator was the same as Glette et al. 2012 \cite{glette}. C++ code for the the Nvidia PhysX physics engine approximately describes the QuadraTot, including the mass and size of the robot components and its degrees of freedom~(\figref{quadratotWhiteBooties}). In the simulator, each individual joint range was limit as described above, but Smart Cropping was not included because it hindered performance by limiting the types of gaits evolution could explore in early generations.
